miniKanren with fair search strategies

### 2019-04-30 Paper/paper.tex
add abstract back; revise revise revise

@@

We propose three new search strategies, balanced interleaving DFS (biDFS), fair 
DFS (fDFS), and BFS. They have different characteristics of fairness 
(Table.~\ref{fairness}). We prove that our BFS and the BFS proposed by Seres et 
al~\citep{seres1999algebra} produce the same result when queried. But our code 
is shorter and runs faster. Because the two BFSs are equivalent, we do not 
distinguish them in most places, except section 6 (quantitative evaluation).

@@

On the contrary, search strategies with fair \disj{} (e.g. fDFS and BFS) give a 
nice answer list.

@@

Search strategies with almost-fair \disj{} (e.g. biDFS) give the same result in 
this case. However, as its name suggests, almost-fair strategies are not always 
fair. Fortunately, the maximal ratio of the allocated resource among 
disjunctive goals is bounded by a constant. Our biDFS is fair when the number 
of goals is a power of 2, otherwise, some goals are allocated twice more 
resource than the others. In following example, the clauses of \texttt{b}, 
\texttt{c}, and \texttt{d} are allocated more resource.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute  $1/4$ of the answer list. However, 
search strategies with unfair \conj{} (e.g. iDFS, fDFS) give us much 
more lists of \texttt{a} than other sorts of lists. And some sorts (e.g. lists 
of \texttt{c}) are hardly found. The situation would be worse if we add more 
\conde clauses. The result with biDFS, whose \conj{} is also unfair, is 
similar, but due to its different \disj{}, the position of \texttt{b} and 
\texttt{c} and swapped.

@@

On the contrary, search strategies with fair \conj{} (e.g. BFS) give a nice 
answer list.

@@

Fairness is good but is not good enough to kick away completeness. A solution  
is bagging states and requiring that 
search spaces derived from states in the same bag are treated fairly. A natural 
way to bag states is by their costs.
This approach is also taken in \citep{seres1999algebra}.
The \emph{cost} of a state is its depth in 
the search tree (i.e. the number of calls to relational definitions required to 
find them) \citep{seres1999algebra}. BFS is more than fair -- it produces 
answers in increasing order of cost. Running the same program gives an answer 
list sorted by answers' costs. In this case, the cost of an answer is equal to 
the length of the inner lists plus the length of the outer list.

Our definition of fair \conj{} is difference from the one in 
\citep{seres1999algebra}. They consider a \conj{} (or "selection rule" in their 
words) is fair if it ``would allow one to choose the literals in different 
order''\citep{seres1999algebra}. This requirement is not stronger than 
\emph{completeness}. We prefer our definition because it avoids overlapping 
with a known concept.

@@

\section{balanced interleaving DFS}

@@

\section{fair DFS}

@@

Our BFS is fair in both \disj{} and \conj{}. Its implementation is based on 
fDFS (not iDFS). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

The implementation can be improved in two aspects. First, as we mentioned in 
section 2.2, our BFS bag states by their cost. However, in this implementation, 
it is unclear where this information is recorded. Second, 
\texttt{append-inf/fair} is space inefficient. It makes $O(n+m)$ new 
\texttt{cons} cells every time, where $n$ and $m$ are the ``length''s of 
input search spaces. We address these issues in the first subsection.

Both our BFS and Seres's BFS \citep{seres1999algebra} produce answers in 
increasing order of cost. So it is interesting to see if they are equivalent. 
We prove the so in Coq. The details are in the second subsection.

@@

As we mentioned in section 2.2, our BFS bag states by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of states in thunk are 
computed recursively then increased by one. It is even more difficult to see 
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make things clear by changing the type of search space, modify 
related function definitions, and introducing a few more functions.

The new type is a pair whose \texttt{car} is a list of state (the bag), and 
whose \texttt{cdr} is either a \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} indicates that the search space is obviously 
finite.

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt}. The 
first three functions in Fig.~\ref{BFS-opt} are search space constructor. 
They are newly introduced. \texttt{none} makes an empty search space. 
\texttt{unit} makes an space from one state. \texttt{step} makes a space from a 
thunk. The remaining functions do the same thing as 
before. Now it should be not difficult to see that \texttt{append-inf/fair} and 
\texttt{append-map-inf} do respect cost information.

Luckily, the change in \texttt{append-inf/fair} also fixes the miserable space 
inefficiency -- the use of \texttt{append} helps us to reuse the first bag of 
\texttt{t-inf}.

Noted that some functions in the list constitute a \emph{MonadPlus}: 
\texttt{none}, \texttt{unit}, \texttt{append-map-inf}, and \texttt{append-inf} 
correspond to \texttt{mzero}, \texttt{unit}, \texttt{bind}, and \texttt{mplus} 
respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, is used to implement \texttt{ifte} and 
\texttt{once}. It takes a space \texttt{s-inf} and two continuations 
\texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some states, 
\texttt{ks} is called with the first state and the rest space. Otherwise, 
\texttt{kf} is called with no argument. Here `s' and `f' means `succeed' and 
`fail' respectively. The remaining functions do the same thing as before.

@@

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code), and embed it into miniKanren. The 
translation is fairly straightforward due to the similarity in both logic 
programming system and search space representation. The translated code is less 
efficient when running our benchmark. Details about efficiency difference are 
in section 6.

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first three benchmarks are taken from 
\citep{friedman_reasoned_2018}. Next two benchmarks about quine are modified 
from a similar test case in \citep{byrd2017unified}. The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and forbids closures going into list. The two benchmarks 
differ in the \conde clause order of their relation interpreters. The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relation 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that iDFS would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. Our statistics support our conjecture.

In general, only iDFS and biDFS constantly perform well. Among them, biDFS 
seems to be less sensitive to change in \conde clause order (see the last four 
benchmarks). The other search strategies all have fair \disj{}. And they all 
perform badly in the \texttt{very-recursiveo} benchmark. However, the drawback 
of having fair \disj{} alone (i.e. fDFS) is not shown elsewhere. Fair \conj{} 
impose overhead constantly except in \texttt{appendo}. The reason might be that 
strategies with fair \conj{} tend to keep more intermediate states in the 
memory. Among the BFSs, our version performs better in most cases, and equally 
well elsewhere.

@@

>>>>

The syntax of a programming language should reflect its semantics. When writing 
a \conde expression in miniKanren, a programmer would expect all clauses share 
the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving DFS (iDFS), however, 
prioritize its clauses by the order how they are written down. Similarly, when 
a \conde is followed by another goal conjunctively, a programmer would expect 
states in parallel share the same chance of being explored. Again, the answers 
by iDFS is different from the expectation. We have devised three new search 
strategies that have different level of fairness in \disj and \conj.

@@

We propose three new search strategies, balanced interleaving DFS (biDFS), fair 
DFS (fDFS), and BFS. They have different characteristics of fairness 
(Table.~\ref{fairness}). We prove that the pure subset of our BFS and the BFS 
proposed by Seres et
al~\citep{seres1999algebra} produce the same result when queried. But our code 
is shorter and runs faster. Because the two BFSs are equivalent, we do not 
distinguish them in most places.

@@

We borrow the definition of \emph{fair} \disj{} from \cite{seres1999algebra}: 
search strategies with \emph{fair} \disj{} should allocate resource evenly 
among disjunctive goals. Running the same program with fDFS and BFS give the 
following result.

@@

We conceive a middle place between fair and unfair -- search strategies with 
\emph{almost-fair} \disj{} should allocate resource so evenly among disjunctive 
goals\citep{seres1999algebra} that the the maximal ratio of resource is bounded 
by a constant. Our new search strategy, biDFS, has almost-fair \disj{}. It is 
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice more resource than the others. In the previous example, it 
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resource.

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each sort constitute $1/4$ of the answer list. However, 
search strategies with unfair \conj{} (e.g. iDFS, fDFS) give us much 
more lists of \texttt{a} than other sorts of lists. And some sorts (e.g. lists 
of \texttt{c}) are hardly found. Although iDFS's \disj{} is unfair in general, 
it is fair when there is no call to relational definition in sub-goals, 
including this case. The situation would be worse if we add more \conde 
clauses. The result with biDFS, whose \conj{} is also unfair, is similar, but 
due to its different \disj{}, the position of \texttt{b} and \texttt{c} are 
swapped.

@@

Intuitively, search strategies with fair \conj{} should produce each sort of 
lists equally frequently. Indeed, BFS does so.

@@

Our solution is requiring a search strategy with \emph{fair} \conj{} to 
package states in bags, where each bag contains finite states, and to allocate 
resource evenly among search spaces derived from states in the same bag. The 
way to package depends on search strategy. And how to allocate resource among 
search space related to different bags is intended left unspecified.

BFS packages states by their costs. The \emph{cost} of a state is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citep{seres1999algebra}. In the following example, the 
cost of an answer is equal to the length of the inner lists plus the length of 
the outer list. Noted that BFS produces answers in increasing order of cost.

@@

\section{balanced interleaving depth-first search}

@@

\section{fair depth-first search}

@@

BFS is fair in both \disj{} and \conj{}. Our implementation is based on 
fDFS (not iDFS). All we have to do is apply two trivial changes to 
\texttt{append-map-inf}. First, rename it to \texttt{append-map-inf/fair}. 
Second, replace its use of \texttt{append-inf} to \texttt{append-inf/fair}.

The implementation can be improved in two aspects. First, as we mentioned in 
section 2.2, our BFS bag states by their cost. However, in this implementation, 
it is unclear where this information is recorded. Second, 
\texttt{append-inf/fair} is extravagant in memory usage. It makes 
$O(n+m)$ new \texttt{cons} cells every time, where $n$ and $m$ are the 
``length''s of input search spaces. We address these issues in the first 
subsection.

Both our BFS and Seres's BFS \citep{seres1999algebra} produce answers in 
increasing order of cost. So it is interesting to see if they are equivalent. 
We prove so in Coq. The details are in the second subsection.

@@

As we mentioned in section 2.2, our BFS bag states by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of states in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make this facts obvious by changing the type of search space, 
modifying related function definitions, and introducing a few more functions.

The new type is a pair whose \texttt{car} is a list of state (the bag), and 
whose \texttt{cdr} is either a \texttt{\#{}f} or a thunk returning a search 
space. A falsy \texttt{cdr} means the search space is obviously finite.

Functions related to the pure subset are listed in Fig.~\ref{BFS-opt} (the 
others in Fig.~\ref{BFS-opt-cont}). The first three functions in 
Fig.~\ref{BFS-opt} are search space constructor. \texttt{none} makes an empty 
search space. 
\texttt{unit} makes an space from one state. \texttt{step} makes a space from a 
thunk. The remaining functions do the same thing as before. Now it should be 
not hard to see that \texttt{append-inf/fair} and \texttt{append-map-inf} do 
respect cost information.

Luckily, the change in \texttt{append-inf/fair} also fixes the miserable space 
extravagance -- the use of \texttt{append} helps us to reuse the first bag of 
\texttt{t-inf}.

Noted that some functions here constitute a \emph{MonadPlus}: 
\texttt{none}, \texttt{unit}, \texttt{append-map-inf}, and \texttt{append-inf} 
correspond to \texttt{mzero}, \texttt{unit}, \texttt{bind}, and \texttt{mplus} 
respectively.

@@

Functions implementing impure features are in Fig.~\ref{BFS-opt-cont}. The 
first function, \texttt{elim}, is used to implement \texttt{ifte} and 
\texttt{once}. It takes a space \texttt{s-inf} and two continuations 
\texttt{ks} and \texttt{kf}. When \texttt{s-inf} contains some states, 
\texttt{ks} is called with the first state and the rest space. Otherwise, 
\texttt{kf} is called with no argument. Here `s' and `f' means `succeed' and 
`fail' respectively. This function is an eliminator of search space, whence the 
name. The remaining functions do the same thing as before.

@@

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code). The translation is fairly straightforward 
due to the similarity in both logic programming system and search space 
representation. The translated code is longer and slower. Details about 
efficiency difference are in section 6.

@@

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first three benchmarks are taken from 
\citep{friedman_reasoned_2018}. Next two benchmarks about quine are modified 
from a similar test case in \citep{byrd2017unified}. The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde clause order of their relation interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relation 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that iDFS would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

In general, only iDFS and biDFS constantly perform well. fDFS is just as 
efficient in all benchmarks but \texttt{very-recursiveo}. Both BFS have obvious 
overhead in many cases. Among the three variants of DFS (they all have unfair 
\conj{}), fDFS is most resistant to clause permutation, followd by biDFS then 
iDFS. Among the two implementation of BFS, ours constantly performs as well or 
better. Interestingly, every strategies with fair \disj{} suffers in 
\texttt{very-recursiveo} and fDFS performs well elsewhere. Therefore, this 
benchmark might be a special case. Fair \conj{} imposes overhead constantly 
except in \texttt{appendo}. The reason might be that strategies with fair 
\conj{} tend to keep more intermediate states in the memory.
